The Two Dualities of Computation: Negative and Fractional Types

### 2012-02-28 rational.tex
organization and abstract

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

@@

\title{Computing in the Field of Rationals}
\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}
\maketitle

\begin{abstract}
  Every functional programmer knows about sum and product types, $a+b$
  and $a*b$ respectively. Negative and fractional types, $a-b$ and
  $a/b$ respectively, are much less known and their computational
  interpretation is unfamiliar and often complicated. We show that in
  a programming model in which information is preserved (such as the
  model introduced in our recent paper on \emph{Information Effects}),
  these types have particularly intuitive and natural computational
  interpretations. Intuitively, values of negative types are values
  that flow ``backwards'' to satisfy demands, values of fractional
  types are values that represent first-class ``structural''
  constraints.
  %% , and values of square root and imaginary types are values that can
  %% be related by arbitrary algebraic constraints.
  The combination of these negative, fractional, square root, and
  imaginary types enable greater flexibility in programming by
  breaking complicated invariants into local ones that can be
  independently satisfied by a subcomputation. Furthermore, they allow
  a programmer to express a wide range of programming idioms,
  including higher-order functions, delimited continuations,
  speculative computation, logic programming, and more.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\category{D.3.1}{Formal Definitions and Theory}{}
\category{F.3.2}{Semantics of Programming Languages}{}
\category{F.3.3}{Studies of Program Constructs}{Type structure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Negative Types: Intuition}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Fractional Types: Intuition}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Summary}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{ {{langRev}} }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Semantics}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{ {{langRevEE}} }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Additive Duality in {{langRev}} }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Multiplicative Duality in {{langRev}} }

@@

\subsection{Syntax}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Rules for {{eta}} and {{eps}} }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions in {{langRevEE}} }
\label{sec:monoidal-constructions}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions Specific to Negation}
\label{sec:neg-constructions}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions Specific to Multiplication}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{To think about}

@@

\subsection{SAT solver}

\subsection{GoI machine}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

@@

\title{The Two Dualities of Computation: Negative and Fractional Types}
\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}
\maketitle

\begin{abstract}
  Every functional programmer knows about sum and product types, {{a+b}} and
  {{a*b}} respectively. Negative and fractional types, $a-b$ and $a/b$
  respectively, are much less known and their computational interpretation is
  unfamiliar and often complicated. We show that in a programming model in
  which information is preserved (such as the model introduced in our recent
  paper on \emph{Information Effects}), these types have particularly
  intuitive and natural computational interpretations. Intuitively, values of
  negative types are values that flow ``backwards'' to satisfy demands and
  values of fractional types are values that impose constraints on their
  context.  The combination of these negative and fractional types enables
  greater flexibility in programming by breaking global invariants into local
  ones that can be autonomously satisfied by a subcomputation. Furthermore,
  they allow a programmer to express a wide range of programming idioms,
  including higher-order functions, delimited continuations, speculative
  computation, logic programming, and more.
\end{abstract}

\category{D.3.1}{Formal Definitions and Theory}{}
\category{F.3.2}{Semantics of Programming Languages}{}
\category{F.3.3}{Studies of Program Constructs}{Type structure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Field of Rational Types: Intuitions}

%%%%%%%%%%%%%%%%%%
\subsection{Negative Types}

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Fractional Types}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Summary}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{ {{langRev}} }

@@

%%%%%%%%%%%%%%%%%%%
\subsection{Semantics}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{ {{langRevEE}} }

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Additive Duality in {{langRev}} }

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection{Multiplicative Duality in {{langRev}} }

@@

%%%%%%%%%%%%%%%%%%
\subsection{Syntax}

@@

%%%%%%%%%%%%%%%%%%%%
\subsection{Rules for {{eta}} and {{eps}} }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions in {{langRevEE}} }
\label{sec:monoidal-constructions}

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions Specific to Negation}
\label{sec:neg-constructions}

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions Specific to Multiplication}

%%%%%%%%%%%%%%%%%%%%%%
\subsection{To think about}

@@

%%%%%%%%%%%%%%%%%%%
\subsection{SAT solver}

%%%%%%%%%%%%%%%%%%%
\subsection{GoI machine}

@@

%% Jacques
